
```{r, output=FALSE, warning=FALSE}
library(tidyverse)
library(tsibble)
library(tsbox)
library(fable)
library(feasts)
library(eurostat)
library(quantmod)
```

# Statistics Review

Let $X$ and $Y$ be two discrete random variables with possible values $0, 1, \ldots, K$. The probability mass functions $p_X(x)$ and $p_Y(y)$ assign occurrence probabilities to each possible outcome.

For example:

```{r, echo=FALSE, warning=FALSE, output=FALSE}
px <- read_csv("https://github.com/febse/data/raw/refs/heads/main/ts/review/px.csv")
py <- read_csv("https://github.com/febse/data/raw/refs/heads/main/ts/review/py.csv")
pxy <- read_csv("https://github.com/febse/data/raw/refs/heads/main/ts/review/pxy.csv") %>%
  select(x, y, p)
```

$$
p_X(x) = \begin{cases}
0.250 & \text{for } x = 0\\
0.095 & \text{for } x = 1 \\
0.272 & \text{for } x = 2 \\
0.383 & \text{for } x = 3 \\
0 & \text{otherwise}
\end{cases}
$$

$$
p_Y(y) = \begin{cases}
0.76 & \text{for } y = 2\\
0.24 & \text{for } y = 3 \\
0 & \text{otherwise}
\end{cases}
$$


For convenience, the probability mass functions of $X$ and $Y$ are stored in the tibbles `px` and `py`.

```{r}
#| label: tbl-distr-x
#| tbl-cap: "Probability mass function of $X$."
 
px %>%
  knitr::kable()
```

```{r}
#| label: tbl-distr-y
#| tbl-cap: "Probability mass function of $Y$"
 
py %>%
  knitr::kable()
```

The probability functions sum to one over all possible values by definition:

$$
\sum_{x = 0}^{3}p_X(x) = 0.250 + 0.095 + 0.272 + 0.383 = 1
$$
```{r}
sum(px$p)
```

:::{#exm-sampling-x}
## Sampling from a univariate distribution

```{r}
set.seed(12)

smpl_x <- px %>%
  slice_sample(n = 1e5, weight_by = p, replace = TRUE)

head(smpl_x) %>%
  knitr::kable(digits=3)
```

It is useful to visualize the observed frequency of each outcome.

```{r}
# Absolute frequencies (counts)
table(smpl_x$x)

# Relative frequencies (shares)
table(smpl_x$x) / nrow(smpl_x)
```


```{r}
smpl_x %>%
  ggplot(aes(x = x)) + 
    geom_bar() + 
    labs(
      x = "Outcome",
      y = "Frequency"
    )
```

:::

:::{#exr-prob-y}
## Probabilities

Show that the probability mass function of $Y$ in @tbl-distr-y sums to one.
Sample 10000 values from its distribution and compute the sample frequency of each outcome. 
:::
:::{.solution}
```{r}
# Type your code here

```
:::

## Expectation

The *expected value* of a random variable is the average of all possible values that can occurr, weighted by their occurrence probabilities. It is a measure of the *location* of the distribution.

$$
\begin{align}
\mu_x & = E(X) = \sum_{x = 0}^{3} x p_X(x) = 0 \times 0.250 + 1 \times 0.095 + 2 \times 0.272 + 3 \times 0.383 = 1.788 \\
\end{align}
$$

```{r}
mu_x <- sum(px$x * px$p)
mu_x
```


:::{#exr-expectation}
## Expected Value

Compute the expected value of $Y$.
:::
:::{.solution}
```{r}
# Type your code here

```
:::


If you want to predict future values of a random variable, the expected value is your best guess in the sense that it minimises the expected value of the quadratic loss function:

$$
E[(X - \hat{x})^2]
$$

Let us construct an example. You need to predict the result of $X$ and you think that the best prediction is $\bar{x} = 1$. When the game runs it will produce four possible values: 0, 1, 2, and 3. The error that you will make is:

$$
L(x) = (x - 1)^2 =
\begin{cases}
   (0 - 1)^2 = 1 & \text{x = 0}\\
   (1 - 1)^2 = 0 & \text{x = 1}\\
   (2 - 1)^2 = 1 & \text{x = 2}\\
   (3 - 1)^2 = 4 & \text{x = 3}
\end{cases}
$$

:::{#exr-expected-loss}
## Expected Quadratic Loss

Compute the expected quadratic loss for a prediction $\bar{x} = 1.5$.

:::
:::{.solution}
```{r}
## Type your code here

# px %>%
#   mutate(
#     loss = ?
#   )
```

:::


## Variance

The variance of a random variable measures how different the possible values that can occur are. Values that occur more often (have higher probability) under $p_X$ receive a higher weight. Values that occur less frequently under $p_X$ are given a lower weight in the sum.

$$
\begin{align}
Var(X) & = \sum_{x = 0}^{3} (x - \mu_x)^2 \times p_X(x) \\
       & = (0 - 1.788)^2 \times 0.250 + (1 - 1.788)^2 \times 0.095 + (2 - 1.788)^2\times 0.272 + (3 - 1.788)^2 \times 0.383 \\
       & = (-1.788)^2 \times 0.250 + (-0.788)^2 \times 0.095 + (0.212)^2\times 0.272 + (1.212)^2 \times 0.383 \\
       & = 3.196 \times 0.250 + 0.620^2 \times 0.095 + 0.044 \times 0.272 + 1.468 \times 0.383 \\
       & \approx 1.433
\end{align}
$${#eq-variance}

```{r}
(px$x - mu_x)
(px$x - mu_x)^2
px$p * (px$x - mu_x)^2
sum(px$p * (px$x - mu_x)^2)
```


You can see from @eq-variance that it is the expected value of the squared diviations from the expected value.

$$
Var(X) = E(X - E(X))^2
$$

:::{#def-variance}

## Variance
The variance of a random variable (distribution) is a summary of the distribution and 
describes its spread: how different are the values that this distribution will generate.

$$
Var(X) = E[(X - E(X))^2] = E(X^2) - E(X)^2
$$
:::

:::{#exr-variance}
## Variance

Compute the Variance of $Y$.
:::
:::{.solution}

```{r}
# Type your code here

```

:::


:::{#thm-exp-value-props}
## Properties of the Expectation

Let $X$ be a random variable with expected value $E(X)$, let $Y$ be a random variable with expected value $E(Y)$, and let $a$ be a fixed constant ($a \in \mathbb{R}$). The following properties are true:

$$
\begin{align}
E(a) & = a \\
E(aX) & = aE(X) \\
E(X + Y) & = E(X) + E(Y)
\end{align}
$$

Furthermore, if $X$ and $Y$ are *uncorrelated*, then the expected value of the product of the two random variables equals the product of their expected values:

$$
E(XY) = E(X)E(Y)
$$
:::

:::{#thm-variance-short}
## Properties of the Variance

Let $X$ be a random variable with expected value $E(X)$, let $Y$ be a random variable with expected value $E(Y)$, and let $a$ be a fixed constant ($a \in \mathbb{R}$). The following properties are true:

$$
Var(X) = E(X^2) - E(X)^2
$$

$$
\begin{align}
Var(a) & = 0 \\
Var(aX) & = a^2 Var(X)
\end{align}
$$
Furthermore, if $X$ and $Y$ are *uncorrelated*, then the variance of their sum equals the sum of their variances:

$$
Var(X + Y) = Var(X) + Var(Y)
$$

:::



:::{#exr-sum}
## Expected value and variance

Use the distributions of $X$ and $Y$ from @tbl-distr-x and @tbl-distr-y to compute the expected value and the variance of

$$
2X + 3Y + 1.
$$

Assume that $X$ and $Y$ are independent.
:::
:::{.solution}
$$
E(2X + 3Y + 1) = \\
Var(2X + 3Y + 1) = 
$$
:::


## Joint Distribution

It is as summary of the *joint* distribution of $X$ and $Y$. The joint probability mass function tells you the probability of the *simultaneous* occurrence of $x$ and $y$. For example, you can ask it the question: what is the probability of $x = 2$ *and* $y = 3$.

For two discrete variables, it is convenient to present the joint distribution as a table with cell entries holding the probabilities. The joint distribution is given in the tibble `pxy` in a long format.

```{r}
pxy %>%
  knitr::kable()
```

Sometimes it is more convenient to see this distribution in a wide format:

```{r}
#| label: tbl-distr-dep
#| tbl-cap: "Joint distribution of $X$ and $Y$."

pxy %>%
  pivot_wider(
    id_cols = x,
    names_from = y,
    values_from = p,
    names_prefix="y="
    ) %>%
  knitr::kable(digits = 3)
```

$$
p_{XY}(x=2, y=3) = 0.043
$$

The joint probability distribution function must sum (integrate) to one over all possible *pairs* of $x$ and $y$.

$$
\sum_{x = 0}^{3}\sum_{y = 2}^{3} p_{XY}(x, y) = 1
$$

```{r}
sum(pxy$p)
```

In the example until now we have summarized the *marginal* distributions of $X$ and $Y$ but we have said nothing about their joint distribution. Usually the joint distribution is determined by the subject matter at hand, but for the sake of example we will look at two joint distributions so that we can get an idea how they work.

First we will construct a special joint distribution under the assumption of *independence*. Intuitively, two random variables are independent, if the outcome of one of the variables does not influence the probability distribution of the other. Imagine that you hold two lottery tickets: one from a lottery in Germany and another from a lottery in Bulgaria. It would be safe to assume that the realized winnings from the German lottery will not affect the odds to win from the Bulgarian ticket.

Now let us consider a case of *dependent* random variables. Let $X$ be the level of a river (at some measurement point) at time $t$ and $Y$ be the level of the same river five minutes later. It woule be safe to assume that if the level of the river was high at $t$ this would affect the distribution of the level of the river at $t$ plus five minutes.

## Marginal Distributions

The marginal distribution of $X$ is obtained by summing the joint distribution of $X$ and $Y$ over all possible values of $Y$.

$$
p_X(x) = \sum_{y=2}^{3}p_{XY}(x, y)
$$

## Conditional Distributions

```{r}
#| label: tbl-cond-distr-y-dep-x
#| tbl-cap: "Conditional distributions of $Y$ given $X$"

pxy_w <- pxy %>%
  pivot_wider(
    id_cols = x,
    names_from = y,
    values_from = p,
    names_prefix = "y="
  ) %>%
  mutate(
    p_x = `y=2` + `y=3`,
    `y=2` = `y=2` / p_x,
    `y=3` = `y=3` / p_x
  )

pxy_w %>%
  knitr::kable(digits = 3)
```

Looking at the conditional distributions of $Y$ given $X$ in @tbl-cond-distr-y-dep-x, you should notice that these are not the same for each value of $X$. For example, $Y=2$ is much more likely when $X = 0$ compared to $X = 3$.



## Joint Distribution under Independence

Lets construct the joint distribution $p_{XY}(x, y)$ that assigns a probability to the points $(x, y)$, assuming that $X$ and $Y$ are independent.

For independent random variables the joint probability of occurrence is simply the product of the *marginal* distributions.

$$
p_{XY}(x, y) = p_X(x)p_Y(y)
$$

```{r}
pxy_ind <- expand_grid(
  px %>% rename(p_x = p), 
  py %>% rename(p_y = p)
)
pxy_ind <- pxy_ind %>%
  mutate(
    p = p_x * p_y
  )
```

```{r, collapse=TRUE}
#| label: tbl-distr-indep
#| tbl-cap: "Joint distribution of $X$ and $Y$ under independence."

pxy_ind_w <- pxy_ind %>%
  pivot_wider(
    id_cols = x,
    names_from = y,
    values_from = p,
    names_prefix = "y="
  )

pxy_ind_w %>%
  knitr::kable(digits = 3)
```
Let's look at the *conditional* distributions of $Y$ given $X$. These answer the questions of the type: if $X$ turns out to be $0$, what are the probabilities for $Y = 2$ and $Y = 3$.

To get the conditional distributions of Y for each possible value of $X$ we divide the cells of the joint distribution table by the marginal probabilities of each $x$.

$$
p_{Y|X}(x, y) = \frac{p_{XY}(x, y)}{p_X(x)}
$$


```{r}
#| label: tbl-xy-distr-ind-w
#| tbl-cap: "Conditional distributions of $Y$. Independence case."

pxy_ind_w %>%
  mutate(
    p_x = `y=2` + `y=3`,
    `y=2` = `y=2` / p_x,
    `y=3` = `y=3` / p_x
  ) %>%
  knitr::kable(digits = 3)
```

What you should see in @tbl-xy-distr-ind-w is that the conditional distributions of $Y$ are the same for every possible value of $X$. This is of course a consequence of the way we constructed this joint distribution in the first place: namely, we assumed that $X$ and $Y$ are independent.



## Conditional Expectation

We have seen how we derived the conditional distributions of $Y$ given $X$ in the previous section. Now we can ask the question: what is the expected value of $Y$ given that $X$ has already turned out to be 0 (for example). We can take the conditional distribution of $Y$ given $X = 0$ and compute the expected value of this distribution.

For the joint distribution under independence:

$$
E(Y | X=0) = \sum_{y = 2}^{3} y p_{Y|X=0}(y) = 2 \times 0.76 + 3 \times 0.24 = 2.24
$$

```{r}
2 * 0.76 + 3 * 0.24
```

For the joint distribution in @tbl-cond-distr-y-dep-x the conditional expectation of $Y$ given $X = 0$ is

$$
E(Y | X=0) = \sum_{y = 2}^{3} y p_{Y|X=0}(y) = 2 \times 0.964 + 3 \times 0.036 = 2.036
$$

```{r}
2 * 0.964 + 3 * 0.036
```

Let us write the conditional expectation of $Y$ for each possible value of $X$ for the dependent joint distribution case.

$$
E(Y | X = x) = \begin{cases}
  2.036 & \text{for } x = 0 \\
  2.060 & \text{for } x = 1 \\
  2.158 & \text{for } x = 2 \\
  2.475 & \text{for } x = 3
\end{cases}
$$

```{r}
#| label: tbl-cond-exp-y
#| tbl-cap: "Conditional expectation of $Y$ for each possible value of $X$."
#| 
pxy %>%
  group_by(x) %>%
  summarise(
    y = y,
    p_y_x = p / sum(p)
  ) %>%
  summarise(
    E_Y_given_X = sum(y * p_y_x) 
  ) %>%
  knitr::kable(digits = 3)
```

An important thing to see here is that the conditional expectation is different for each value of $X$. As the value of $X$ is uncertain (it is a random variable), the conditional expectation of $Y$ given $X$ is also a random variable. Its distribution is given by the possible values and the probabilities of occurrence of $X$ (the marginal distribution of $X$).


:::{#exr-cond-exp-ind}
Calculate the expected value of $Y$ given $X$ for every possible value of $X$ in the case joint distribution under independence.
:::


:::{#exm-sampling-joint}
## Sampling from the Joint Distribution

```{r}
sample_joint <- pxy %>%
  slice_sample(n = 1000, weight_by = p, replace = TRUE)

head(sample_joint)
```

```{r}
sample_joint %>%
  group_by(x, y) %>%
  summarise(
    p = first(p),
    n = n(),
    f = n / nrow(sample_joint)
  )
```
:::

## Covariance

The *covariance* measures the (linear) dependency between two random variables.

:::{#def-covariance}
## Covariance
The covariance of two random variables $X$ and $Y$ is given by

$$
Cov(X, Y) = E[(X - E(X))(Y - E(Y))]
$$
Alternatively, it can be computed using the decomposition formula:

$$
Cov(X, Y) = E(XY) - E(X)E(Y)
$$

:::

In the analysis of time series we will often encounter situations where the expected value of one of the random variables is zero. As can be seen from the decomposition formula, in that case the covariance reduces to

$$
Cov(X, Y) = E(XY).
$$


Closely related to the covariance is the correlation between $X$ and $Y$.

:::{#def-correlation}
## Correlation

$$
\rho(X, Y) = \frac{Cov(X, Y)}{\sqrt{Var(X)Var(Y)}}
$$
It is easy to show that the correlation is bounded between -1 and 1.

$$
-1 \leq \rho(X, Y) \leq 1
$$

:::

:::{#exr-corr-proof}
## Correlation

Let X be a random variable with, and $Y = a + bX$. Show that the correlation between $X$ and $Y$ equals one or minus one depending on the sign of $b$. For simplicity, assume that $E(X) = 0$.


:::



:::{#thm-covariance-props}
## Properties of the Covariance

Let $X$ and $Y$ be random variables and let $a, b \in \mathbb{R}$ be fixed constants.

$$
Var(aX + bY) = a^2 Var(X) + b^2Var(Y) + 2abCov(X, Y)
$$
:::

:::{#exr-covariance}
## Covariance
Compute the covariance of $X$ and $Y$ under the joint distributions given in @tbl-distr-indep and @tbl-distr-dep. Use the `pxy` and `pxy_ind` tables for these calculations.
:::
:::{.solution}
```{r}
# Type your code here

```
:::

:::{#exr-covariance-corr}
## Variance of Correlated Variables
Compute the variance of $2X - Y$ using the same distributions as in @exr-covariance.

:::


## Empirical Estimation of Moments


Throughout this course we will focus on three summaries of the stochastic
processes: the mean (expected value, level), the variance (fluctuation), and the
covariances/correlations (dependency) between lags of the random process.

In the the previous statistics-related courses we had a sample of (uncorrelated) observations
$x_1, x_2,\ldots, x_n$. We assumed that these are samples (realizations) from 
some normal distribution $N(\mu, \sigma^2)$ with unknown mean and variance. We used the observed
values to learn something about this distribution. We estimated its mean with

$$
\hat{\mu} = \frac{1}{n}\sum_{i = 1}^{n} x_i
$$

and its variance with:

$$
\hat{\sigma}^2 = \frac{1}{n  -1}\sum_{i = 1}^{n}(x_i - \hat{\mu})^2
$${#eq-evar}

In the case of time series, however, we have only _one_ observation for each time point.


```{r, warning=FALSE}
n <- 10
B <- 5
set.seed(21)

dt_rw <- tibble(
  r = rep(1:B, n),
  e = rnorm(n = B*n)
) %>%
  group_by(r) %>%
  mutate(
    t = 1:n,
    x = cumsum(e),
    is_observed = (r == 1),
  )

dt_rw %>%
  ggplot(
    aes(
      x = t,
      y = x, 
      group = r, 
      alpha = factor(is_observed))
    ) +
  geom_line() + 
  geom_point() + 
  labs(
    alpha = "Observed",
    y = expression(x[t])
  ) + 
  scale_x_continuous(breaks = 1:n)
```

In order to be able to apply our usual technique for moments estimation
we have to _assume_ that the random process is ergodic, meaning that we can
estimate its statistical properties (mean, variance, correlations) from the
its time series (observed values). We will talk more extensively about (weak) _stationarity_, 
which is a necessary (though not sufficient) condition for ergodicity. For (weakly) stationary time series the 
mean, variance and covariances do not change over time.

:::{#def-autocorrelations}
## Autocovariances and Autocorrelations

$$
\begin{align}
\mu & = E(X_1) = E(X_2) = \ldots = E(X_T) \\
\gamma(0) & = Var(X_1) = Var(X_2) = \ldots = Var(X_T) \\
\gamma(1) & = Cov(X_1, X_2) = Cov(X_2, X_3) = \ldots = Cov(X_{t - 1}, X_{t}) = Cov(X_{T - 1}, X_{T}) \\
\gamma(k) & = Cov(X_{t}, X_{t + k}) \\
\rho(k) & = \frac{\gamma(k)}{\gamma(0)} 
\end{align}
$$

:::

For a random process $X_1, X_2,\ldots, X_T$ and an observed time series $x_1, x_2, \ldots, x_T$, we will estimate the mean with the arithmetic average of the
observed values.

$$
\hat{\mu} = \frac{1}{T}\sum_{t = 1}^{T} x_{t}
$$

We will estimate the variance of a stationary process with the _empirical_ variance:

$$
\hat{\gamma}(0) = \frac{1}{T}\sum_{t = 1}^{T}(x_t - \bar{x})^2
$${#eq-ts-evar}

And the covariances with the _empirical_ (auto-)covariances:

$$
\hat{\gamma}(k) = \frac{1}{T}\sum_{t = 1}^{T - k} (x_{t} - \hat{\mu})(x_{t + k} - \hat{\mu})
$${#eq-ts-ecov}

and empirical (auto-)correlations:

$$
\hat{\rho}(k) = \frac{\hat{\gamma}(k)}{\hat{\gamma}(0)}
$$
To see the meaning of the autocovariances, let us create two artificial time
series and plot their lags. The time series will be generated from two processes.
The first one will be called the _purely random_ or _white noise_ process.

$$
e_t \sim N(0, 1) \text{ for every } t \\
E(e_t) = 0 \\
Var(e_t) = 1 \\
Cov(e_t, e_{t + k}) = 0, k \neq 0
$$
The second one is be called a _random walk_.

$$
y_t = \sum_{i = 1}^{t} e_i
$$


```{r}
set.seed(432)

# Create 20 observations from two time series. rnorm draws random numbers
# from a normal distribution with zero expected value (mean) and standard
# deviation (sd) equal to 1. We call this distribution the standard normal
# distribution. In the mutate step we create the first two lags of the series

dt_sim <- tibble(
  x = rnorm(n = 20, mean = 0, sd = 1),
  y = cumsum(x)
) %>%
  mutate(
    x_l1 = lag(x, n = 1),
    y_l1 = lag(y, n = 1)
  )

mean_x <- mean(dt_sim$x)
mean_x

mean_y <- mean(dt_sim$y)
mean_y
```


```{r}
dt_sim %>%
  ggplot(aes(x = x, y = x_l1)) +
  geom_point() + 
  geom_vline(xintercept=mean_x, lty=2, alpha = 0.5) + 
  geom_hline(yintercept=mean_x, lty=2, alpha = 0.5) + 
  labs(
    x = "x",
    y = "lag(x, 1)"
  )
```

Looking at the plot, think about the signs of the cross-products that go into
the formula for the covariance:

$$
(x_{t + 1} - \hat{\mu})(x_{t} - \hat{\mu})
$$

Where do you find positive and negative cross-products in the plot?

```{r}
dt_sim %>%
  ggplot(aes(x = y_l1, y = y)) +
  geom_point() + 
  geom_vline(xintercept=mean_y, lty=2, alpha = 0.5) + 
  geom_hline(yintercept=mean_y, lty=2, alpha = 0.5) + 
  labs(
    x = "lag(y, 1)",
    y = "y"
  )
```


The following code computes the covariance 
```{r}
# Compute the covariance
cov(dt_sim$x, dt_sim$x_l1, use = "complete.obs")

# COmpute the correlation
cor(dt_sim$x, dt_sim$x_l1, use = "complete.obs")
```

Note that `cov` computes the covariance according to formula @eq-evar.


## Exercises

The following code downloads 

```{r download-bitcoin}
bitcoin <- getSymbols(
  "BTC-USD", 
  from="2015-01-01",
  to="2023-01-01",
  src="yahoo", 
  periodicity="daily", 
  env = NULL)

names(bitcoin) <- c("Open", "High", "Low", "Close", "Volume", "Adjusted")

bc <- bitcoin %>%
  ts_tsibble() %>%
  pivot_wider(
    names_from = id,
    values_from = value
  )
```


```{r}
# Plots the closing exchange rate against the time index

bc %>%
  autoplot(Close) + 
  labs(
    # Set the labels for the x and y axis
    x = "Date",
    y = "USD/BTC"
  )
```

1. Compute the average exchange rate (`Close`).
```{r}
# Type your code here

```

2. Compute the empirical variance using the `var` function

```{r}
# Type your code here

```


3. Construct a two new variables that contain the first and the second lags of `Close`. Leave only `Close` and its lags in the resulting tsibble. Hint: Use the `mutate` and `lag` functions and select the columns using `select`.

```{r}
# Type your code here
# Select the code below and press Ctrl-Shift-c in order to remove the comments


```

4. Create a scatterplot for `Close` and its first lag. Hint: Use the code
generating the scatterplot in the simulation above and adapt it.

```{r}
# Type your code here

```

5. Compute the covariance between `Close` and its first lag. Compute the autocorrelation coefficient (empirical autocorrelation) between `Close` and its first lag. Hint: Look into the documentation of `cov` and `cor` to find out how to handle missing values.

```{r}
# Type your code here

```

6. Use the `gg_lag` function to create scatterplots for the between the time series
and its first six lags. What patterns can you see in these scatterplots?

```{r}
bc %>%
  gg_lag(
    Close, 
    lags = 1:6, 
    geom="point", 
    size = 0.5
  )
```

7. Use the `acf` function to plot the autocorrelation function of the `Close` time series.

```{r}
# Type your code here

```



